---
title: "Fill Missing Phenology Dates"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
dat = read_csv("Data/final_metric_files/final_combined_dates.csv")
lakes = c("AL", "BM", "CB", "CR", "SP", "TB", "TR", "FI", "ME", "MO", "WI")
dat$lakeid = factor(dat$lakeid, 
                    levels = rev(lakes), 
                    ordered = T)
vars = c("iceoff", "straton", "secchi_all", "secchi_openwater", "daphnia_biomass",
         "total_zoop_biomass", "chlor_all", "chlor_spring", "chlor_fall", "doc_epiMax",
         "totpuf_epiMax", "totpuf_epiMin", "totpuf_hypoMax", "totpuf_hypoMin", 
         "anoxia_summer", "stability", "energy", "stratoff", "iceon")
dat$metric = factor(dat$metric,
                    levels = rev(vars),
                    ordered=T)
```
# Data

## Which years are missing a metric
```{r, fig.width = 14, fig.height=20}
dat %>% 
  ggplot(aes(year, lakeid, fill=is.na(daynum))) +
  geom_tile(color="grey") +
  theme_bw() +
  facet_wrap(~metric, nrow=6) 
  # labs(fill = "Missing") + 
  # scale_fill_scico(palette = 'imola')

# dat %>% 
#   ggplot(aes(year, lakeid, fill=daynum)) +
#   geom_tile(color="grey") +
#   theme_bw() +
#   facet_wrap(~metric, nrow=6) + 
#   # labs(fill = "Missing") + 
#   scale_fill_viridis_c()
```

**Remaining data issues:**

* Northern lakes don't have DOC until until 1986; missing first 4 years when all other variables are available. Potential fixes:
  - Exclude DOC; start analyses in 1986; fill DOC with e.g., median value, or **model of other variables**?
* Two missing iceon dates in AL (1982) and CB (1988). Potential fix:
  - **Fill with function/model of other lake ice on dates**
* Southern lakes missing chlorophyll data: 1996 - 1999 and 2002 - 2004: is this the fluorometer issue? **See if other instruments (spec) have data to use for these years.**
* Zoop data issues:
  - CB missing daphnia peak 1995(?): **fill with medain**
  - FI missing several years from 1996 - 2006; unsure what to do.
  
  
**Predict ice-on in northern lakes from other lakes ice-on dates**  
```{r}
n_iceoff_wide = dat %>% 
  select(-sampledate)%>% 
  filter(lakeid %in% c("AL", "BM", "CB", "CR", "SP", "TB", "TR") & metric == "iceoff") %>% 
  pivot_wider(names_from = lakeid, values_from = daynum)

al_iceoff_model = lm(AL~BM+CB+CR+SP+TB+TR, data=n_iceoff_wide)
summary(al_iceoff_model)
plot(n_iceoff_wide$AL, predict(al_iceoff_model, n_iceoff_wide), pch=16)
# actually looks good; use to fill
cb_iceoff_model = lm(CB~AL+BM+CR+SP+TB+TR, data=n_iceoff_wide)
summary(cb_iceoff_model)
plot(n_iceoff_wide$CB, predict(cb_iceoff_model, n_iceoff_wide), pch=16)
# really good; go ahead and fill

al_missing_iceoff = n_iceoff_wide %>% filter(is.na(AL))
al_missing_iceoff$iceoff_fill = round(predict(al_iceoff_model, al_missing_iceoff))
al_iceoff_fill = al_missing_iceoff %>% 
  select(metric, year, iceoff_fill) %>% 
  rename(daynum_fill = iceoff_fill) %>% 
  mutate(metric = "iceoff", lakeid = "AL") 

cb_missing_iceoff = n_iceoff_wide %>% filter(is.na(CB))
cb_missing_iceoff$iceoff_fill = round(predict(cb_iceoff_model, cb_missing_iceoff))
cb_iceoff_fill = cb_missing_iceoff %>% 
  select(metric, year, iceoff_fill) %>% 
  rename(daynum_fill = iceoff_fill) %>% 
  mutate(metric = "iceoff", lakeid = "CB")

iceoff_fill = bind_rows(al_iceoff_fill, cb_iceoff_fill)
```

  
**Predict DOC daynum from other variable daynum's in each northern lake**
```{r}
no_dups = dat %>% group_by(lakeid, year, metric) %>% summarise(N = n()) %>% filter(N == 1)

n_lakes_wide = left_join(no_dups, dat) %>% 
  select(-sampledate) %>% 
  pivot_wider(names_from = metric, values_from = daynum) %>% 
  filter(lakeid %in% c("AL", "BM", "CB", "CR", "SP", "TB", "TR"))

hold_data = na.omit(data.frame(n_lakes_wide %>%  ungroup() %>% select(-year, -N))) # , -doc_predict
all = lm(doc_epiMax~(iceon+straton+stratoff+energy+
                 stability+anoxia_summer+secchi_all+
                 secchi_openwater+total_zoop_biomass+daphnia_biomass)*lakeid, data=hold_data)
i_o = lm(doc_epiMax~1, data=hold_data)
hold_step = step(i_o, scope=formula(all))

doc_model = lm(doc_epiMax~(iceon+straton+stratoff+energy+
                 stability+anoxia_summer+secchi_all+
                 secchi_openwater+total_zoop_biomass+daphnia_biomass)*lakeid, 
               data=n_lakes_wide, 
               na.action = na.exclude)
summary(doc_model)

n_lakes_wide$doc_predict = predict(doc_model, n_lakes_wide)

cor = n_lakes_wide %>% 
  group_by(lakeid) %>% 
  summarise(r = paste("r =", round(cor(doc_epiMax, doc_predict, use="complete.obs"), 3)))

n_lakes_wide %>% 
  ggplot(aes(x=doc_epiMax, y=doc_predict, color=as.character(lakeid))) +
  geom_point()+
  theme_bw() +
  labs(color="lakeid")+
  geom_abline(slope=1, intercept = 0) + 
  facet_wrap(~lakeid) + 
  geom_text(aes(label=r), data=cor, x=300, y=0) +
  labs(x="obs peak DOC (daynum)",  y="modeled peak DOC (daynum)")

# not good but vaguely positive relationship; use this instead of median?
missing_doc = n_lakes_wide %>% filter(is.na(doc_epiMax))

missing_doc$doc_fill = round(predict(doc_model, missing_doc))

doc_fill = missing_doc %>% 
  select(lakeid, year, doc_fill) %>% 
  rename(daynum_fill = doc_fill) %>% 
  mutate(metric = "doc_epiMax") %>% 
  filter(year < 2000)


```

## Try to fill the missing chlorophyll values in the southern lakes

**Using dates from fitting wtemp/do model to chl observations
```{r}
filled_chl_dates = read_csv("Data/derived/chl_filled_peaks.csv") %>% 
  rename(daynum_fill = daynum) %>% 
  select(-sampledate)
```


## Create filled metric column and filled
```{r, fig.width = 14, fig.height=14}
all_fill = bind_rows(iceoff_fill, doc_fill, filled_chl_dates)
dat_filled = full_join(dat, all_fill) %>% 
  mutate(filled_flag = ifelse(is.na(daynum) & !is.na(daynum_fill), TRUE, FALSE)) %>% 
  mutate(daynum_fill = ifelse(is.na(daynum) & !is.na(daynum_fill), daynum_fill, daynum))

dat_filled$lakeid = factor(dat_filled$lakeid, 
                    levels = c("AL", "BM", "CB", "CR", "SP", "TB", "TR", "FI", "ME", "MO", "WI"), 
                    ordered = T)
vars = c("iceoff", "straton", "secchi_all", "secchi_openwater", "daphnia_biomass",
        "total_zoop_biomass", "chlor_all", "chlor_spring", "chlor_fall", "doc_epiMax",
        "totpuf_epiMax", "totpuf_epiMin", "totpuf_hypoMax", "totpuf_hypoMin", 
        "anoxia", "anoxia_summer", "stability", "energy", "stratoff", "iceon")
dat_filled$metric = factor(dat_filled$metric,
                    levels = rev(vars),
                    ordered=T)

dat_filled %>% 
  ggplot(aes(year, metric, fill=is.na(daynum_fill))) +
  geom_tile(color="grey") +
  theme_bw() +
  facet_wrap(~lakeid, nrow=6) # + 
  # labs(fill = "Missing") + 
  # scale_fill_viridis_c() 
```
## Additional trimming/filling

Trim to "good" years, fill FI ice data with Monona, then fill additional missing with median values for each lake/metric
```{r, fig.width = 14, fig.height=14}
df_yearsWant = dat_filled %>% 
  filter((lakeid %in% c("FI", "ME", "MO", "WI") & year %in% 1996:2018) |
           (!(lakeid %in% c("FI", "ME", "MO", "WI")) & year %in% 1982:2018))

fi_icefill = df_yearsWant %>% 
  filter(lakeid == "MO" & metric %in% c("iceoff", "iceon")) %>% 
  mutate(lakeid = "FI") %>% 
  select(-daynum, -filled_flag, -sampledate) %>% 
  rename(icefill = daynum_fill)

df_yearsWant = df_yearsWant %>% 
  full_join(fi_icefill) %>% 
  mutate(daynum_fill = ifelse(is.na(daynum_fill) & !is.na(icefill), 
                              icefill, daynum),
         filled_flag = ifelse(is.na(daynum) & !is.na(daynum_fill) & !is.na(icefill), 
                              TRUE, filled_flag))

df_yearsWant %>% 
  ggplot(aes(year, metric, fill=filled_flag)) +
  geom_tile(color="grey") +
  theme_bw() +
  facet_wrap(~lakeid, nrow=6)
```

### Final fill w/ medians
```{r}
all_lys = df_yearsWant %>% 
  select(lakeid, year) %>% 
  distinct()

metric = unique(df_yearsWant$metric)

all_lys_want = expand_grid(all_lys, metric)

dat_final = left_join(all_lys_want, df_yearsWant)

medians = dat_final %>% 
  group_by(lakeid, metric) %>% 
  summarise(median_daynum = median(daynum_fill, na.rm=T))

dat_final = dat_final %>% 
  left_join(medians) %>% 
  mutate(daynum_fill = ifelse(is.na(daynum_fill), median_daynum, daynum_fill)) %>% 
  mutate(filled_flag = ifelse(is.na(daynum) & !is.na(daynum_fill), TRUE, filled_flag)) %>% 
  select(-icefill, -median_daynum)
```

```{r, fig.width = 14, fig.height=14}
dat_final %>%  
  ggplot(aes(year, metric, fill=daynum_fill)) +
  geom_tile(color="grey") +
  theme_bw() +
  facet_wrap(~lakeid, nrow=6)

write_csv(dat_final, "Data/analysis_ready/final_combined_dates_filled_v2.csv")
```



=========================================================================================
## Old code: no longer used

### chlor_all
```{r}
no_dups = dat %>% group_by(lakeid, year, metric) %>% summarise(N = n()) %>% filter(N == 1)

s_lakes_wide = left_join(no_dups, dat) %>% 
  select(-sampledate) %>% 
  pivot_wider(names_from = metric, values_from = daynum) %>% 
  filter(lakeid %in% c("FI", "ME", "MO", "WI"))

# stepwise
hold_data = na.omit(data.frame(s_lakes_wide %>% filter(lakeid != "FI") %>% select(-chlor_spring, -chlor_fall, -daphnia_biomass, -total_zoop_biomass)))
all = lm(chlor_all~., data=hold_data)
i_o = lm(chlor_all~1, data=hold_data)
hold = step(i_o, scope=formula(all))

chlAll_model_MEMOWI = lm(chlor_all~anoxia_summer + lakeid + iceon + stability + straton + stratoff, 
               data=s_lakes_wide %>% filter(lakeid != "FI"))
summary(chlAll_model_MEMOWI)
plot(s_lakes_wide %>% filter(lakeid != "FI") %>% pull(chlor_all), 
     predict(chlAll_model_MEMOWI, s_lakes_wide %>% filter(lakeid != "FI")),
     col=s_lakes_wide$lakeid, pch=16, xlab="observed", ylab="predicted")


# using to fill for now
missing_chlAll_MEMOWI = s_lakes_wide %>% filter(is.na(chlor_all) & lakeid != "FI")
missing_chlAll_MEMOWI$chl_all_fill = round(predict(chlAll_model_MEMOWI, missing_chlAll_MEMOWI))

chlAll_model_FI = lm(chlor_all~(stratoff+energy+
                 stability+anoxia_summer+secchi_all+
                 secchi_openwater), 
               data=s_lakes_wide %>% filter(lakeid == "FI"))
summary(chlAll_model_FI)
plot(s_lakes_wide %>% filter(lakeid == "FI") %>% pull(chlor_all), 
     predict(chlAll_model_FI, s_lakes_wide %>% filter(lakeid == "FI")),
     col=s_lakes_wide$lakeid, pch=16)

missing_chlAll_FI = s_lakes_wide %>% filter(is.na(chlor_all) & lakeid == "FI")
missing_chlAll_FI$chl_all_fill = round(predict(chlAll_model_FI, missing_chlAll_FI))
# gives negative value; DONT fill FI

chlAll_fill = missing_chlAll_MEMOWI %>% 
  select(lakeid, year, chl_all_fill) %>% 
  rename(daynum_fill = chl_all_fill) %>% 
  mutate(metric = "chlor_all") 
```

### chlor_spring
```{r}
# stepwise
hold_data = na.omit(data.frame(s_lakes_wide %>% filter(lakeid != "FI") %>% select(-chlor_all, -chlor_fall, -daphnia_biomass, -total_zoop_biomass)))
all = lm(chlor_spring~., data=hold_data)
i_o = lm(chlor_spring~1, data=hold_data)
hold = step(i_o, scope=formula(all))

chlSpring_model_MEMOWI = lm(chlor_all~stability + straton + doc, 
               data=s_lakes_wide %>% filter(lakeid != "FI"))
summary(chlSpring_model_MEMOWI)
plot(s_lakes_wide %>% filter(lakeid != "FI") %>% pull(chlor_spring), 
     predict(chlSpring_model_MEMOWI, s_lakes_wide %>% filter(lakeid != "FI")),
     col=s_lakes_wide$lakeid, pch=16)


# not great; don't fill it
# missing_chlSpring_MEMOWI = s_lakes_wide %>% filter(is.na(chlor_spring) & lakeid != "FI")
# missing_chlSpring_MEMOWI$chl_all_fill = round(predict(missing_chlSpring_MEMOWI, missing_chlSpring_MEMOWI))
# 
# # gives negative value; DONT fill FI
# 
# chlAll_fill = missing_chlSpring_MEMOWI %>% 
#   select(lakeid, year, chl_all_fill) %>% 
#   rename(daynum_fill = chl_all_fill) %>% 
#   mutate(metric = "chlor_spring") 
```
### chlor_fall
```{r}
# stepwise
hold_data = na.omit(data.frame(s_lakes_wide %>% filter(lakeid != "FI") %>% select(-chlor_all, -chlor_spring, -daphnia_biomass, -total_zoop_biomass)))
all = lm(chlor_fall~., data=hold_data)
i_o = lm(chlor_fall~1, data=hold_data)
hold = step(i_o, scope=formula(all))

chlFall_model_MEMOWI = lm(chlor_all~iceoff + year + lakeid, 
               data=s_lakes_wide %>% filter(lakeid != "FI"))
summary(chlFall_model_MEMOWI)
plot(s_lakes_wide %>% filter(lakeid != "FI") %>% pull(chlor_fall), 
     predict(chlFall_model_MEMOWI, s_lakes_wide %>% filter(lakeid != "FI")),
     col=s_lakes_wide$lakeid, pch=16)
# not great; don't fill it

# missing_chlSpring_MEMOWI = s_lakes_wide %>% filter(is.na(chlor_spring) & lakeid != "FI")
# missing_chlSpring_MEMOWI$chl_all_fill = round(predict(missing_chlSpring_MEMOWI, missing_chlSpring_MEMOWI))
# 
# # gives negative value; DONT fill FI
# 
# chlAll_fill = missing_chlSpring_MEMOWI %>% 
#   select(lakeid, year, chl_all_fill) %>% 
#   rename(daynum_fill = chl_all_fill) %>% 
#   mutate(metric = "chlor_spring") 
```

